Appendix S1. Linear regression models and non-linear population dynamics

### 2024-05-03 Appendix_S1_v01.Rmd
Env lm: quadratic term added

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

@@

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# model function
disc_log_E <- function(r, N0, alpha, E, x) {
    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
p1 + p2
```

@@

```{r,warning=FALSE}
# Fit the model
m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + E[1:(t-1)])
#plotting the series along with the fitted values
m.2.ar.fit <- log(N) - residuals(m.2.ar)
m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
dat$ar2.fit <- m.2.ar.fit
dat$lm2.fit <- NA
dat$lm2.fit[2:t] <- m.2.lm.fit
```
```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```
The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ are no longer well-represented by a linear function without taking $E$ into account.

@@

d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
d.dN$E <- dat$E[2:t]
p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()

@@

@@

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- data.frame(x=cbind(log(dat$N[1:(t-1)])),E[1:(t-1)])
colnames(XData)[2] <- "E"
m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x+E)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m2.post.hmsc <- convertToCodaObject(m.2.sample)
summary(m2.post.hmsc$Beta)
plot(m2.post.hmsc$Beta)
```

@@

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.2.ar$coef
m.2.ar$coef[2]*(1-m.2.ar$coef[1])
# linear model
summary(m.2.lm)$coefficients[1:3,1:2]
# Bayesian estimates
summary(m2.post.hmsc$Beta)$statistics[1:3,1:2]
```

@@

```{r figs9, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(x=list(2)),ngrid=39)
predY <- predict(m.2.sample,Gradient=Gradient,expected=TRUE)
#preds <-computePredictedValues(m.1.sample)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
```

@@

```{r}
blah <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)]))
colnames(blah) <- c("Nt1","Nt","E")
mod <- lm(Nt1 ~ Nt + E,data=blah)
jtools::effect_plot(mod,pred=E,plot.points = TRUE)
```

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

```{r}
lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
colnames(lm_dat) <- c("Nt1","Nt")
ggplot(lm_dat,aes(Nt,Nt1)) +
  stat_summary(fun.data= mean_cl_normal) + 
  geom_smooth(method='lm')
```

@@

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# model function
disc_log_E <- function(r, N0, alpha, E, x) {
    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
p1 + p2
```

@@

```{r,warning=FALSE}
# Fit the model
m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + log(E[1:(t-1)]))
#plotting the series along with the fitted values
m.2.ar.fit <- log(N) - residuals(m.2.ar)
m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
dat$ar2.fit <- m.2.ar.fit
dat$lm2.fit <- NA
dat$lm2.fit[2:t] <- m.2.lm.fit
```
```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

@@

d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
d.dN$E <- log(dat$E[2:t])
p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()

@@

The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ can still be captured by a linear relationship. However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear. This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable). The use of linear relationships in JSDMs is discussed in [@Ingram2020], and in many applications (e.g. [@Erickson2023]) quadratic terms are used, which create bell-shaped response curves that may better match species with optimal niches (as opposed to linear, monotonically increasing relationships between population size and environmental predictors). We thus include a quadratic term for $E_t$ to provide a better fit to the data.

```{r}
df <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2))
colnames(df) <- c("Nt1","Nt","E","Esq")
m.2.lm <- lm(Nt1 ~ Nt + E + Esq,data=df)
```

```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
ggplot(data = blah, aes(x = E, y = Nt1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
```

@@

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- df
m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m2.post.hmsc <- convertToCodaObject(m.2.sample)
summary(m2.post.hmsc$Beta)
bayesplot::mcmc_trace(m2.post.hmsc$Beta)
bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
```

@@

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.2.ar$coef
m.2.ar$coef[2]*(1-m.2.ar$coef[1])
# linear model
summary(m.2.lm)$coefficients[1:4,1:2]
# Bayesian estimates
summary(m2.post.hmsc$Beta)$statistics[1:4,1:2]
```

@@

```{r figs9, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
# Can't figure out curved gradient
pr <- predict(m.2.hmsc,m.2.sample,)
```

@@
